---
title: "Assignment 1: Data Set Selection and Initial Processing"
author: "Priyanka Narasimhan"
output: html
---

# This R Notebook will provide insight on the initial preparation and analysis of the dataset GSE23118. 
# Resources: Sample code from lecture slides, and RPR-GEO2R.R from Prof. Streipe.

# Install and load additional packages like BiocManager, BioBase, & GEOquery
```{r}
rm(list = ls())
if (! requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (! requireNamespace("Biobase", quietly = TRUE)) {
  BiocManager::install("Biobase")
}
if (! requireNamespace("GEOquery", quietly = TRUE)) {
  BiocManager::install("GEOquery")
}
# Load
library(BiocGenerics)
library(Biobase)
library(GEOquery)
```

# 1 - Download the dataset from GEO
```{r}
gset <- GEOquery::getGEO("GSE23118", GSEMatrix =TRUE)
length(gset) # the length is 1
edata <- gset[[1]]
edata # This is our expression matrix
str(edata)
```

# Our phenodata returns a dataframe that shows our 16 GEO samples, and 40 variable labels 
```{r}
names = (pData(edata))
names
# We also see 2 supplementary files 
supfiles = getGEOSuppFiles('GSE23118')
fnames = rownames(sfiles)
fnames
```

# 2 - Assessing the Dataset 
# This particular dataset studies the interaction between phage mimetic nanorods and SKBT-3 breast cancer cells. 
# Some general information about this platform:

```{r}
gse <- getGEO("GSE23118", GSEMatrix=FALSE)
current_gpl <- names(GPLList(gse))[1]
current_gpl_info <- Meta(getGEO(current_gpl))
```
**Platform title:** `r current_gpl_info$title`
**Submission date:** `r current_gpl_info$submission_date`
**Last updated date:** `r current_gpl_info$last_update_date`
**Organism:** `r current_gpl_info$organism`
**Number of GEO datasets that use this technology:** `r length(current_gpl_info$series_id)`
**Number of GEO samples that use this technology:** `r length(current_gpl_info$sample_id)`

# I was particularly interested in this dataset because I recently watched a W5 series on phages and
# phage treatments, which are becoming widely popular in helping to cure disorders, bacterial infections # and other immune-related illnesses. This particular experiment uses phage mimetic nanorods as
# gene/drug carriers to specifically-targeted cancer cells.

# Experimental Design:
# Phage-mimetic nanorods (PMNs) were used with 2 different types of cells, SKBR-3 breast cancer cells
# and H441 lung adenocarcinoma cells. Gene expression profiling of cells exposed to PMNs was conducted,
# and the cellular responses were recorded. Two groups of cells were compared, a control group
# (without PMNs) and a tested group. 





